---
title: "GelMedix Data Analysis"
output: html_notebook
author:
  - Tom Wright^[University of Toronto, twright@kensingtonhealth.org]
---

```{r setup, echo=FALSE, warn=FALSE}
library(tidyverse)
library(magrittr)
library(ggplot2)
library(forcats)
library(stringr)
library(drc)
```

```{r load_data, echo=FALSE, warn=FALSE}
data <- read.csv('data/markers.csv')

data <- data %>%
  mutate(Step = factor(Step, levels = c('DA 0.001 cd.s/m2 + OP',
                                        'DA 0.003 cd.s/m2 + OP',
                                        'DA 0.01 cd.s/m2 + OP',
                                        'DA 0.03 cd.s/m2 + OP',
                                        'DA 0.1 cd.s/m2 + OP',
                                        'DA 0.3 cd.s/m2 + OP',
                                        'DA 1 cd.s/m2 + OP',
                                        'DA 3 cd.s/m2 + OP',
                                        'DA 10 cd.s/m2 + OP',
                                        'CW 150 cd/m2',
                                        'LA 0.03 cd.s/m2 + OP',
                                        'LA 0.3 cd.s/m2 + OP',
                                        'LA 3 cd.s/m2 + OP',
                                        'LA 3 cd.s/m2 10 Hz Flicker',
                                        'LA 3 cd.s/m2 20 Hz Flicker',
                                        'LA 3 cd.s/m2 30 Hz Flicker',
                                        'LA 3 cd.s/m2 40 Hz Flicker'))) %>%
  mutate(Timepoint = factor(Timepoint, levels = c('Baseline',
                                                  'Day 8',
                                                  'Day 15',
                                                  'Day 22',
                                                  'Day 30',
                                                  'Day 38'))) %>%
    mutate(days = as.numeric(ifelse(Timepoint=="Baseline", 0, str_extract(Timepoint, "\\d+"))))

data_flash <- data %>% 
  filter(!str_detect(Step, "Flicker")) %>%
  filter(!str_detect(Step, "CW")) %>%
  mutate(Step = fct_drop(Step))

data_flash %<>%
  dplyr::select(!Chan) %>%
  pivot_wider(names_from="Marker", values_from=c("Amp", "Time")) %>%
  mutate('Amp_ab' = Amp_b - Amp_a) %>%
  pivot_longer(cols=starts_with(c('Amp_','Time_'))) %>%
  pivot_wider(names_from="name", values_from="value")

data_flicker <- data %>% 
  filter(str_detect(Step, "Flicker")) %>%
  mutate(Step = fct_drop(Step)) %>%
  dplyr::select(!Time) %>% # drop time values
  pivot_wider(names_from="Marker", values_from="Amp") %>%
  mutate(Amp_1 = Peak1 - Trough1,
         Amp_2 = Peak2 - Trough2,
         Amp_3 = Peak3 - Trough3,
         Amp_4 = Peak4 - Trough4,
         Amp_5 = Peak5 - Trough5) 

 data_flicker <- mutate(data_flicker, Amp = rowMeans(dplyr::select(data_flicker, starts_with("Amp_"))))%>%
   dplyr::select(Subject, Group, Timepoint, Step, Eye, Amp, days) %>%
   mutate(Subject = factor(Subject),
          Group = factor(Group))



data %>%
  dplyr::select(Subject, Group, Timepoint) %>%
  unique() %>%
  group_by(Group, Subject) %>%
  summarise(count=n())

```

# Flash data

```{r}
# Boxplot visualisation of ERG amplitude vs time after intervention by group
data_flash %>% 
  filter(Step %in% c('DA 0.01 cd.s/m2 + OP', 'DA 3 cd.s/m2 + OP', 'LA 3 cd.s/m2 + OP')) %>%
  ggplot(aes(x=Timepoint, y=Amp_ab, color=Group)) +
  geom_boxplot() +
  facet_grid(rows=vars(Step), cols=vars(Eye), scales="free_y")
```
## DA0.01 data only
This data is selected as it's a widely used standard flash intensity

```{r}
# Single variate ANOVA analysis of DA0.01 data

model <- data_flash %>% 
  filter(Step == 'DA 0.01 cd.s/m2 + OP') %>%
  lm(Amp_ab ~ I(Subject) + Group * days - 1, data=.)

anova(model)
summary(model)
  
```
```{r}
# Visualisation of DA0.01 data with linear smoothing
data_flash %>% 
  filter(Step == 'DA 0.01 cd.s/m2 + OP') %>%
  ggplot(aes(x=days, y=Amp_b, color=Group, group=Group)) +
  geom_point() +
  stat_smooth(method=lm, se=TRUE) +
  ggtitle('DA 0.01 Smooth=Linear')
```

```{r}
# Visualisation of DA0.01 data with loess smoothing
data_flash %>% 
  filter(Step == 'DA 0.01 cd.s/m2 + OP') %>%
  ggplot(aes(x=days, y=Amp_ab, color=Group, group=Group)) +
  geom_point() +
  stat_smooth(method=loess, se=TRUE) +
  ggtitle('DA 0.01 Smooth=Loess')
```

```{r}
data_flash %>% 
  filter(Step == 'DA 0.01 cd.s/m2 + OP') %>%
  ggplot(aes(x=Timepoint, y=Amp_ab, fill=Group)) +
  stat_summary(fun="mean", geom="bar", position = position_dodge(0.95)) +
  stat_summary(fun='mean', geom='point', position = position_dodge(0.95), size=1) +
  stat_summary(fun.data = "mean_cl_normal", geom="errorbar", position = position_dodge(0.95), width=0.2)
```

## DA3.0 data only
This data is selected as it's a widely used standard flash intensity

```{r}
model <- data_flash %>% 
  filter(Step == 'DA 3 cd.s/m2 + OP') %>%
  lm(Amp_ab ~ I(Subject) + Group * days - 1, data=.)

anova(model)
summary(model)
  
```
```{r}
data_flash %>% 
  filter(Step == 'DA 3 cd.s/m2 + OP') %>%
  ggplot(aes(x=days, y=Amp_ab, color=Group, group=Group)) +
  geom_point() +
  stat_smooth(method=lm, se=TRUE) +
  ggtitle('DA 3 Smooth=Linear')
```
```{r}
data_flash %>% 
  filter(Step == 'DA 3 cd.s/m2 + OP') %>%
  ggplot(aes(x=days, y=Amp_ab, color=Group, group=Group)) +
  geom_point() +
  stat_smooth(method=loess, se=TRUE) +
  ggtitle('DA 3 Smooth=Loess')
```
## LA3.0 data only
This data is selected as it's a widely used standard flash intensity

```{r}
model <- data_flash %>% 
  filter(Step == 'LA 3 cd.s/m2 + OP') %>%
  lm(Amp_ab ~ I(Subject) + Group * days -1, data=.)

anova(model)
summary(model)
  
```
```{r}
data_flash %>% 
  filter(Step == 'LA 3 cd.s/m2 + OP') %>%
  ggplot(aes(x=days, y=Amp_ab, color=Group, group=Group)) +
  geom_point() +
  stat_smooth(method=lm, se=TRUE) +
  ggtitle('LA 3 Smooth=Linear')
```


```{r}
data_flash %>% 
  filter(Step == 'LA 3 cd.s/m2 + OP') %>%
  ggplot(aes(x=days, y=Amp_ab, color=Group, group=Group)) +
  geom_point() +
  stat_smooth(method=loess, se=TRUE) +
  ggtitle('LA 3 Smooth=Loess')
```
# Light adapted flicker
## 10 Hz
```{r}
data_flicker %>% 
  filter(Step == 'LA 3 cd.s/m2 10 Hz Flicker') %>%
  ggplot(aes(x=days, y=Amp, color=Group, group=Group)) +
  geom_point() +
  stat_smooth(method=lm, se=TRUE) +
  ggtitle('LA Flicker 10Hz. Smooth=Linear')

```
```{r}
model <- data_flicker %>% 
  filter(Step == 'LA 3 cd.s/m2 10 Hz Flicker') %>%
  lm(Amp ~ I(Subject) + Group * days -1, data=.)

anova(model)
summary(model)
  
```
## 20 Hz
```{r}
data_flicker %>% 
  filter(Step == 'LA 3 cd.s/m2 20 Hz Flicker') %>%
  ggplot(aes(x=days, y=Amp, color=Group, group=Group)) +
  geom_point() +
  stat_smooth(method=lm, se=TRUE) +
  ggtitle('LA Flicker 20Hz. Smooth=Linear')

```
```{r}
model <- data_flicker %>% 
  filter(Step == 'LA 3 cd.s/m2 20 Hz Flicker') %>%
  lm(Amp ~ I(Subject) + Group * days -1, data=.)

anova(model)
summary(model)
```
## 30 Hz
```{r}
data_flicker %>% 
  filter(Step == 'LA 3 cd.s/m2 30 Hz Flicker') %>%
  ggplot(aes(x=days, y=Amp, color=Group, group=Group)) +
  geom_point() +
  stat_smooth(method=lm, se=TRUE) +
  ggtitle('LA Flicker 30Hz. Smooth=Linear')

```
```{r}
model <- data_flicker %>% 
  filter(Step == 'LA 3 cd.s/m2 30 Hz Flicker') %>%
  lm(Amp ~ I(Subject) + Group * days -1, data=.)

anova(model)
summary(model)
```
## 40 Hz
```{r}
data_flicker %>% 
  filter(Step == 'LA 3 cd.s/m2 40 Hz Flicker') %>%
  ggplot(aes(x=days, y=Amp, color=Group, group=Group)) +
  geom_point() +
  stat_smooth(method=lm, se=TRUE) +
  ggtitle('LA Flicker 40Hz. Smooth=Linear')

```
```{r}
model <- data_flicker %>% 
  filter(Step == 'LA 3 cd.s/m2 40 Hz Flicker') %>%
  lm(Amp ~ I(Subject) + Group * days -1, data=.)

anova(model)
summary(model)
  
```

# Intensity response functions

Change in b-wave amplitude with increasing stimulus intensity can be described using a Michaelis-Menton hyperbolic function.

__Vmax__ is interpreted as an indicator of photoreceptor / retinal health. A reduction in Vmax can occur secondary to loss of photoreceptors, disruption of the dark current or inner retinal dyfunction. __K__ is interpreted as an index of retinal sensitivity representing the efficiency of quantal capture. [link](https://iscev.wildapricot.org/resources/Documents/StandardsEtc/DarkAdaptedERGb-wave-transfer(intensity-response)function[draft].pdf)

```{r naka_rushton, echo=FALSE, results='asis'}
t <- data_flash %>%
  mutate(Intensity = as.numeric(str_extract(Step, "\\d+\\.*\\d*")))

m<-t %>%
  filter(Condition=='DA' & Timepoint=='Baseline') %>%
  drm(Amp_ab~Intensity, data=., Subject, fct=MM.2(),lowerl=c(0,0), upperl=c(500,1))

plot(m, col=TRUE, lwd=1)

m2<-t %>%
  filter(Condition=='DA' & Timepoint=='Day 8') %>%
  drm(Amp_b~Intensity, data=., Subject, fct=MM.2(),lowerl=c(0,0), upperl=c(500,1))

plot(m, col=TRUE, lwd=1)
```  